---
title: "ADS_503_Diabetes_EDA"
format: html
editor: visual
---

```{r}
library(tidyverse)

diabetes <- read_csv("datasets/diabetic_data.csv")
```

```{r}
dim(diabetes)

names(diabetes)
```

```{r}
summary(diabetes)
```

```{r}
colSums(diabetes == "?", na.rm = TRUE)

diabetes[diabetes == "?"] <- NA
```

```{r}
#eval correlated variable
numeric_vars <- diabetes %>%
  select(where(is.numeric))

cor_matrix <- cor(numeric_vars, use = "pairwise.complete.obs")
```

```{r}
table(diabetes$readmitted)

prop.table(table(diabetes$readmitted))
```

```{r}
ggplot(diabetes,
       aes(x = readmitted)) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Distribution of Readmission Status",
    x = "Readmission Status",
    y = "Count"
  )
```

```{r}
# age vs readmission
ggplot(diabetes,
       aes(x = age,
           fill = readmitted)) +
  geom_bar(position = "fill") +
  labs(
    title = "Readmission Proportion by Age Group",
    x = "Age Group",
    y = "Proportion"
  )
```

```{r}
#time in hospital
ggplot(diabetes,
       aes(x = time_in_hospital)) +
  geom_histogram(bins = 20,
                 fill = "steelblue")
```

```{r}
# time in hospital vs readmission
ggplot(diabetes,
       aes(x = readmitted,
           y = time_in_hospital)) +
  geom_boxplot(fill = "lightblue")
```

```{r}
# number of meds vs readmission
ggplot(diabetes,
       aes(x = readmitted,
           y = number_inpatient)) +
  geom_boxplot(fill = "lightpink")
```

```{r}
# number of inpatiant visits vs readmission 
ggplot(diabetes,
       aes(x = readmitted,
           y = number_inpatient)) +
  geom_boxplot(fill = "lightpink")
```

```{r}
ggplot(diabetes,
       aes(x = time_in_hospital)) +
  geom_histogram(bins = 20,
                 fill = "steelblue")
```

```{r}
# replace all ? with N/A
diabetes[diabetes == "?"] <- NA
```

```{r}
# missing value percentages
missing_pct <- colSums(is.na(diabetes)) /
               nrow(diabetes) * 100
```

```{r}
# remove identifiers
diabetes_clean <- diabetes %>%
  select(-encounter_id, -patient_nbr)
```

```{r}
# change into categorical variables
diabetes_clean <- diabetes_clean %>%
  mutate(across(where(is.character), as.factor))
```

```{r}
#correlation analysis
numeric_vars <- diabetes_clean %>%
  select(where(is.numeric))

cor_matrix <- cor(numeric_vars,
                  use = "pairwise.complete.obs")
```

```{r}
#outlier eval
ggplot(diabetes_clean,
       aes(y = time_in_hospital)) +
  geom_boxplot()
```

# Data Wrangling and Pre-processing

> Covers pre-processing, data splitting, and model strategy.

```{r}
#| label: setup-modeling
#| message: false
#| warning: false
library(tidyverse)
library(caret)      

set.seed(503)
```

The raw `readmitted` column has three levels (`<30`, `>30`, `NO`). We frame the predictive task around the clinically and financially important question of **early (within–30-day) readmission**, which is the outcome hospitals are penalized for under CMS readmission programs. We therefore binarize the target: `readmitted = "<30"` → **Yes** (event), all else → **No**.

**Key data-quality decisions**:

- **Patient-level leakage.** The 101,766 rows are hospital *encounters* for only 71,518 unique patients. Keeping multiple encounters per patient would leak information across the train/test boundary, so we retain one (the first) encounter per patient.
- **Outcome-impossible encounters.** Encounters discharged to death or hospice (`discharge_disposition_id` ∈ {11, 13, 14, 19, 20, 21}) cannot result in readmission and would bias the model; these are removed.
- **Excessive missingness.** `weight` is \~97% missing, `payer_code` \~40% (administrative), and `medical_specialty` \~49% — all are dropped. The small amount of missing `race` (\~2%) is retained as an explicit `"Unknown"` level.
- **High-cardinality diagnoses.** `diag_1/2/3` hold 700+ raw ICD-9 codes each. We collapse them into the standard clinical categories (Circulatory, Respiratory, Diabetes, etc.).
- **Identifiers** (`encounter_id`, `patient_nbr`) and three `Unknown/Invalid` gender rows are removed.

### Feature engineering and cleaning

```{r}
#| label: preprocess

group_icd9 <- function(code) {
  num <- suppressWarnings(as.numeric(code))
  case_when(
    is.na(code)                                    ~ "Other",
    str_starts(code, "250")                        ~ "Diabetes",
    !is.na(num) & ((num>=390 & num<=459)|num==785) ~ "Circulatory",
    !is.na(num) & ((num>=460 & num<=519)|num==786) ~ "Respiratory",
    !is.na(num) & ((num>=520 & num<=579)|num==787) ~ "Digestive",
    !is.na(num) & (num>=800 & num<=999)            ~ "Injury",
    !is.na(num) & (num>=710 & num<=739)            ~ "Musculoskeletal",
    !is.na(num) & ((num>=580 & num<=629)|num==788) ~ "Genitourinary",
    !is.na(num) & (num>=140 & num<=239)            ~ "Neoplasms",
    TRUE                                           ~ "Other"
  )
}

diabetes_model <- diabetes %>%
  mutate(across(everything(), ~ ifelse(.x == "?", NA, .x))) %>% 
  arrange(encounter_id) %>% distinct(patient_nbr, .keep_all = TRUE) %>% 
  filter(!discharge_disposition_id %in% c(11,13,14,19,20,21)) %>%     
  filter(gender != "Unknown/Invalid") %>%                             
  select(-encounter_id, -patient_nbr, -weight, -payer_code, -medical_specialty) %>%
  mutate(across(c(diag_1, diag_2, diag_3), group_icd9)) %>%           
  mutate(race = replace_na(race, "Unknown")) %>%                      
  mutate(readmitted = factor(ifelse(readmitted == "<30", "Yes", "No"),
                             levels = c("Yes", "No"))) %>%            
  mutate(across(c(admission_type_id, discharge_disposition_id,
                  admission_source_id), as.character)) %>%            
  mutate(across(where(is.character), as.factor))

cat("Rows after cleaning:", nrow(diabetes_model),
    "| Predictors:", ncol(diabetes_model) - 1, "\n")
cat("Target (early readmission) balance:\n")
print(round(prop.table(table(diabetes_model$readmitted)), 4))
cat("Remaining NA cells:", sum(is.na(diabetes_model)), "\n")
```

### Near-zero variance and correlated predictors

Many drug columns (e.g., `examide`, `citoglipton`) are essentially constant and carry no signal; we remove near-zero-variance predictors. We then check for redundant numeric predictors via `findCorrelation` (\|r\| \> 0.90).

```{r}
#| label: nzv-corr
nzv <- nearZeroVar(diabetes_model %>% select(-readmitted), names = TRUE)
cat("Near-zero-variance predictors removed (", length(nzv), "):\n  ",
    paste(nzv, collapse = ", "), "\n")
diabetes_model <- diabetes_model %>% select(-all_of(nzv))

num_pred <- diabetes_model %>% select(where(is.numeric))
high_corr <- findCorrelation(cor(num_pred), cutoff = 0.90, names = TRUE)
cat("Highly correlated numeric predictors removed:",
    ifelse(length(high_corr) == 0, "none", paste(high_corr, collapse = ", ")), "\n")
if (length(high_corr) > 0)
  diabetes_model <- diabetes_model %>% select(-all_of(high_corr))

cat("Final modeling frame:", nrow(diabetes_model), "x", ncol(diabetes_model), "\n")
```

# Data Splitting

We use a **stratified 60/20/20 train/validation/test split** on the binarized target so the \~9% early-readmission rate is preserved in every partition. The training set supports cross-validated hyperparameter tuning; the validation set is for model selection across families; the test set is held out for a single, unbiased final evaluation.

```{r}
#| label: split
set.seed(503)
idx_test <- createDataPartition(diabetes_model$readmitted, p = 0.20, list = FALSE)
test_df  <- diabetes_model[idx_test, ]
rest     <- diabetes_model[-idx_test, ]

idx_val  <- createDataPartition(rest$readmitted, p = 0.25, list = FALSE) 
val_df   <- rest[idx_val, ]
train_df <- rest[-idx_val, ]

tibble(
  Set          = c("Train", "Validation", "Test"),
  n            = c(nrow(train_df), nrow(val_df), nrow(test_df)),
  Readmit_rate = round(c(mean(train_df$readmitted == "Yes"),
                         mean(val_df$readmitted   == "Yes"),
                         mean(test_df$readmitted  == "Yes")), 4)
)
```

# Model Strategies

**Primary research question:**

1.  Using demographic, admission, diagnosis, and treatment information captured during a hospital stay, can we predict whether a diabetic patient will be readmitted within 30 days of discharge?
    1.  This is a **binary classification** problem aimed at flagging high-risk patients for follow-up care.

**Secondary questions:**

1.  Which factors are most associated with early readmission (e.g., prior inpatient visits, number of diagnoses, discharge disposition)?
2.  Do flexible non-linear learners outperform interpretable linear baselines on this largely categorical dataset?

**Central challenge: class imbalance** (\~9% positive). Two design choices address it:

- *Resampling:* down-sampling the majority class **within** each cross-validation fold (`sampling = "down"` in `caret::trainControl`), keeping resampled performance estimates honest.
- *Metrics:* accuracy is uninformative (a trivial "no readmission" model scores \~91%). We optimize and report **ROC-AUC** as the primary metric, alongside **sensitivity (recall of readmissions)**, **specificity**, **balanced accuracy**, and **F1**.

All models will be tuned via 5-fold cross-validation on the training set using ROC-AUC, compared on the validation set, and the finalist evaluated once on the test set. The shared resampling scheme is defined below for the model-building stage.

```{r}
#| label: train-control

ctrl <- trainControl(method = "cv", number = 5,
                     classProbs = TRUE,
                     summaryFunction = twoClassSummary,
                     sampling = "down",
                     savePredictions = "final")
```

# Model Building and Hyperparameter Tuning

Each candidate model is tuned with the shared 5-fold cross-validation scheme (`ctrl`) defined above, optimizing ROC-AUC. The additional libraries below support the individual learners.

```{r}
#| label: model-libs
#| message: false
#| warning: false
library(glmnet)  
library(earth)   
library(ranger) 
library(gbm)  
library(pROC)   
```

```{r}
#| label: train-models
#| message: false
#| warning: false

fit <- function(method, ...) {
  train(readmitted ~ ., data = train_df, method = method,
        metric = "ROC", trControl = ctrl, ...)
}

set.seed(503)
m_glm    <- fit("glm", family = "binomial")   
m_glmnet <- fit("glmnet", tuneLength = 5)
m_mars   <- fit("earth", tuneLength = 5)
m_rpart  <- fit("rpart", tuneLength = 8)
m_rf     <- fit("ranger", tuneLength = 4, num.trees = 400, importance = "impurity")
m_gbm    <- fit("gbm", verbose = FALSE,
                tuneGrid = expand.grid(n.trees = c(150, 300, 500),
                                       interaction.depth = c(2, 4, 6),
                                       shrinkage = 0.05, n.minobsinnode = 10))

models <- list(Logistic = m_glm, Penalized = m_glmnet, MARS = m_mars,
               Tree = m_rpart, RandomForest = m_rf, GBM = m_gbm)
```

> Plain logistic regression issues *rank-deficient fit* warnings because several sparse dummy-encoded factor levels are linearly dependent. This is precisely the motivation for the **penalized (glmnet)** model, which regularizes such collinear predictors instead of failing on them.

### Cross-validated model comparison

```{r}
#| label: cv-compare
#| fig-width: 8
#| fig-height: 4
resamps <- resamples(models)
bwplot(resamps, metric = "ROC", main = "Cross-Validated ROC-AUC by Model")
```

# Validation and Testing

Every tuned model is scored on the held-out **validation** set across the imbalance-aware metrics. Because overall accuracy is misleading at \~9% prevalence, we rank models by **ROC-AUC** and inspect sensitivity, specificity, balanced accuracy, and F1. The single best model is then carried to the **test** set for one unbiased estimate of generalization performance.

```{r}
#| label: validation-eval

eval_set <- function(mod, df) {
  p   <- predict(mod, df, type = "prob")[, "Yes"]
  cl  <- predict(mod, df)
  roc <- as.numeric(pROC::auc(pROC::roc(df$readmitted, p,
                     levels = c("No", "Yes"), direction = "<", quiet = TRUE)))
  cm  <- confusionMatrix(cl, df$readmitted, positive = "Yes")
  tibble(ROC_AUC      = roc,
         Sensitivity  = cm$byClass["Sensitivity"],
         Specificity  = cm$byClass["Specificity"],
         Balanced_Acc = cm$byClass["Balanced Accuracy"],
         F1           = cm$byClass["F1"])
}

val_results <- imap_dfr(models, ~ eval_set(.x, val_df) %>% mutate(Model = .y)) %>%
  select(Model, everything()) %>% arrange(desc(ROC_AUC))
knitr::kable(val_results, digits = 3, caption = "Validation-set performance")

best_name <- val_results$Model[1]
best_mod  <- models[[best_name]]
cat("Selected final model:", best_name, "\n")
```

### Final model: variable importance and test performance

```{r}
#| label: final-eval
#| fig-width: 7
#| fig-height: 5
plot(varImp(best_mod), top = 15,
     main = paste("Top 15 Predictors —", best_name))

test_results <- eval_set(best_mod, test_df)
knitr::kable(test_results, digits = 3,
             caption = paste("Test-set performance —", best_name))
```

```{r}
#| label: roc-curve
#| fig-width: 6
#| fig-height: 6
test_prob <- predict(best_mod, test_df, type = "prob")[, "Yes"]
roc_obj   <- pROC::roc(test_df$readmitted, test_prob,
                       levels = c("No", "Yes"), direction = "<", quiet = TRUE)
plot(roc_obj, col = "steelblue", lwd = 2,
     main = paste0(best_name, " — Test ROC (AUC = ", round(pROC::auc(roc_obj), 3), ")"))
```

### Results and final model selection

The penalized logistic regression model was selected as the final model because it had the highest validation ROC-AUC. Its validation ROC-AUC was 0.647, which was only slightly higher than GBM at 0.643 and logistic regression at 0.642. This suggests that several models performed similarly, but the penalized model was the strongest based on the validation-set selection rule.

On the held-out test set, the penalized model had a ROC-AUC of 0.659, with sensitivity of 0.511 and specificity of 0.718. This means the model identified about 51% of true early readmissions while correctly identifying about 72% of non-early-readmission cases. Overall, the model performed better than random guessing, but its predictive strength was still moderate. These results suggest that the model may be useful as a screening or prioritization tool, but it should not be used as a stand-alone clinical decision tool.


